A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-08-02 analysis/paper/paper.Rmd
Minor edits

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no]. It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018]. All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a digital research compendium following @marwick2018.

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure. As the data collection for the employed DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation. In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

>>>>

The elevation data used for the analysis is a digital terrain model (DTM) freely available from the Norwegian Mapping Authority [@mapping2018, https://hoydedata.no]. It was here opted for the 10m resolution DTM rather than the higher-resolution 1m version. In addition to resulting in considerably less processing time, the higher resolution elevation model is more vulnerable to smaller-scale modern disturbances that the 10m version is not impacted by. The 10m resolution DTM of the study area is a down-sampled version of the 1m version and has a height accuracy with a systematic error of 0.1m [@mapping2018]. All data and R programming code [@rcoreteam2021] required to run the analyses, as well as the derived data are freely available in an online repository at <https://osf.io/7f9su/>, organised as a research compendium following @marwick2018.

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure. As the data collection for the employed DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation. In addition to employing the 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.
